---
title: "Validating Models and Traits"
author: "Jamie Prentice"
date: "9 Dec 2021"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(results = "hide")
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r, results = "hide", echo = FALSE}
source("libraries.R")
source("source_files.R")
set.seed(0)
```

# Simulating test data

We simulate an SIDR epidemic with the Fishboost demography and groups. We assume genetic variation in susceptibility, infectivity, and recoverability only, not time to detection.

```{r, results = "hide", cache = TRUE}
params <- make_parameters(
    model_type = "SIDR",
    setup = "fishboost",
    cov = "sir_only",
    group_layout = "fishboost"
)

pedigree <- make_pedigree(params)
traits <- make_traits_from_pedigree(pedigree, params)
traits <- set_groups(traits, params)
pop <- simulate_epidemic(traits, params)
```

Plot the results to check things have worked okay:

```{r, results = "hide", echo = FALSE, fig.width = 7, out.width = "100%"}
plt <- plot_model(pop, params)
```

## Validation

### Test times ~ traits

We test if phenotypic recoverability matches the time $t(T_\text{sym} \to T_\text{rec})$. This would be a huge cause for concern if it didn't match.

```{r}
with(pop, cor.test(Trec - Tsym, recoverability_ln,
                   method = "spearman"))

with(pop, cor.test(Trec - Tsym, recoverability_ln,
                   method = "pearson"))
```

I tried Anova too, just in case.

```{r}
anova(lm(Trec - Tsym ~ recoverability_ln, data = pop))
```

Do the same for susceptibility, although this time we remove the donors, since their susceptibility didn't matter.

```{r}
pop2 <- pop[donor == 0]
with(pop2, cor.test(Tinf, susceptibility_ln,
                   method = "spearman"))

with(pop2, cor.test(Tinf, susceptibility_ln,
                    method = "pearson"))

anova(lm(Tinf ~ susceptibility_ln, data = pop2))
```

## Check heritability

We check that the offspring traits correlate with that of their parents by taking the mean trait values for the progeny of each sire / dam.

```{r}
traitnames    <- c("susceptibility", "infectivity", "recoverability")
traitnames_BV <- paste0(traitnames, "_BV")
traitnames_m  <- paste0(traitnames, "_m")

sire_BVs <- pop[sdp == "sire", ..traitnames_BV]
dam_BVs  <- pop[sdp == "dam",  ..traitnames_BV]

for (sire_id in pop[sdp == "sire", id]) {
    sire_means <- pop[sire == sire_id,
                      lapply(.SD, mean),
                      .SDcols = traitnames_BV]
    pop[id == sire_id, (traitnames_m)] <- sire_means
}

for (dam_id in pop[sdp == "dam", id]) {
    dam_means <- pop[dam == dam_id,
                     lapply(.SD, mean),
                     .SDcols = traitnames_BV]
    pop[id == dam_id, (traitnames_m)] <- dam_means
}

with(pop, cor.test(susceptibility_m, susceptibility_BV))
with(pop, cor.test(infectivity_m, infectivity_BV))
with(pop, cor.test(recoverability_m, recoverability_BV))
```

Time to infection and recovery periods depend _indirectly_ on the susceptibility and recovery traits, so visually check that these match up, along with the parent-offspring trait correlations. Note that SIRE needs to go from $T_\text{sym}$ / $T_\text{rec}$ to offspring traits to parent traits, so that's twice separated.

```{r, results = "hide", echo = FALSE, fig.dim = c(12, 8), out.width="100%"}
library(cowplot)

plt_TTI <- ggplot(pop2, aes(x = susceptibility, y = Tinf)) +
    geom_point() + geom_smooth(method = "lm") +
    labs(x = "Susceptibility", y = "Time to infection (days)")

plt_RP <- ggplot(pop, aes(x = recoverability, y = Trec - Tsym)) +
    geom_point() + geom_smooth(method = "lm") +
    labs(x = "Recoverability", y = "Recovery period (days)")
    

plt_sus <- ggplot(pop, aes(x = susceptibility_m, y = susceptibility_BV)) +
    geom_point() + geom_smooth(method = "lm") +
    labs(x = "Mean progeny susceptibility", y = "Parent susceptibility")
plt_inf <- ggplot(pop, aes(x = infectivity_m, y = infectivity_BV)) +
    geom_point() + geom_smooth(method = "lm") +
    labs(x = "Mean progeny infectivity", y = "Parent infectivity")
plt_rec <- ggplot(pop, aes(x = recoverability_m, y = recoverability_BV)) +
    geom_point() + geom_smooth(method = "lm") +
    labs(x = "Mean progeny recoverability", y = "Parent recoverability")

plot_grid(plotlist = list(plt_sus, plt_inf, plt_rec, plt_TTI, plt_RP))
```
