---
title: "Full Pipeline"
author: "Jamie Prentice"
date: '2021-03-02'
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

## Overview

Create simulated epidemic data and send to SIRE 2.0. The intention is to replace simulated data with Turbot data.

R functions based on Richard Bailey's originals. [SIRE 2.0](https://github.com/theITEAM/SIRE2.0/) modified from Chris Pooley's original.


## Load Libraries and source files
```{r include = FALSE}
library(data.table) # used for all data table manipulations
library(Matrix)     # allows Matrix
library(xml2)       # used to generate XML file
# library(corpcor)    # efficient estimation of Cov and (partial) Corr
library(MASS)       # needed for mvnorm
# library(MCMCglmm)
library(ggplot2)    # plotting
library(tictoc)     # check performance
library(codetools)  # check for global variables
library(here)       # useful according to Hadley Wickham

source("make_parameters.R")
source("get_R0.R")
source("make_pedigree.R")
source("make_grm.R")
source("make_traits_from_grm.R")
source("make_traits_from_pedigree.R")
source("set_groups.R")
source("simulate_epidemic.R")
source("init_populations.R")
source("prepare_data.R")
source("data_table_to_tsv_string.R")
source("generate_sire_xml.R")
source("make_plots.R")
source("utils.R")
# source("sire_data.R")
```

## Generate pedigree, GRM, and traits
Given a set of initial parameters, first generate a pedigree with the required structure, and from that generate the corresponding *A matrix*. Then generate a population with appropriate traits, either using the pedigree or the A matrix (since we may want to be able to use a GRM later).
```{r}
    # params <- make_parameters(epi_type = "SEIR", name = "seir1", cov = TRUE, setup = "")
    params <- make_parameters(epi_type = "SIR", name = "sir1", cov = TRUE, setup = "")

    pedigree <- make_pedigree(params)
    # this is an A-matrix, not actually a GRM
    GRM <- make_grm(pedigree)

    # traits <- make_traits_from_grm(GRM, pedigree, params)
    traits <- make_traits_from_pedigree(pedigree, params)

    # check_cov(traits2, params)
```

## Simulate an epidemic and plot the results
```{r}
traits <- set_groups(traits, params)
pop    <- simulate_epidemic(traits, params)
get_R0(pop)

# Plot the epidemic ----
plot_model(pop, params)
```

## Generate the XML file for SIRE 2.0
```{r}
# Discretised version of pop (Tinf, Trec -> "[S,1][I,2][R,3]")
# dpop <- discretise(pop, params)


# pop is offspring only, so need to recombine parent data
# Tidy up ready for saving data to XML file
# Handy to have tmax stored here, because it's harder to get afterwards
data_and_tmax <- prepare_data(pop, params)


# Generate the XML file for SIRE 2.0
generate_sire_xml(data_and_tmax, GRM, params)

# copy the XML file to the SIRE 2.0 folder to run
system(paste("cp", params$filename, "../sire2/"))

## Run SIRE 2.0 ----
cmd <- paste0("../sire2/sire ", params$filename, " 0")
message("Running SIRE 2.0 ...")

# message(paste0("> system(\"", cmd, "\")"))
system(cmd)

# In case we need to test if the output is unchanged from the unmodified SIRE 2.0
# system("../SIRE2.0/Execute/sire foo.xml 0")
```

